### 2017-04-27 paper/jfp-submission/abstract.tex
tweak numbers...

\begin{abstract}
  %
  {Static} type errors are a common stumbling block
  for newcomers to typed functional languages.
  %
  We present a {dynamic} approach to explaining type
  errors by generating counterexample witness inputs that
  illustrate how an ill-typed program goes wrong.
  %
  First, given an ill-typed function, we symbolically
  execute the body to synthesize witness values that
  make the program go wrong.
  We prove that our procedure synthesizes
  {general} witnesses in that if a witness is
  found, then for all inhabited input types,
  there exist values that can make the function go wrong.
  %
  Second, we show how to extend the above procedure to
  produce a reduction graph that can be used to
  interactively visualize and debug witness executions.
  %
  Third, we evaluate the coverage of our approach
  on two data sets comprising over 4,500 ill-typed
  student programs.
  Our technique is able to generate witnesses for
  88\% of the programs, and our reduction graph
  yields small counterexamples for 81\% of the witnesses.
  %
  Finally, we evaluate whether our witnesses help
  students understand and fix type errors, and
  find that students presented with our witnesses
  show a greater understanding of type errors
  than those presented with a standard error message.
\end{abstract}

>>>>

\begin{abstract}
  %
  {Static} type errors are a common stumbling block
  for newcomers to typed functional languages.
  %
  We present a {dynamic} approach to explaining type
  errors by generating counterexample witness inputs that
  illustrate how an ill-typed program goes wrong.
  %
  First, given an ill-typed function, we symbolically
  execute the body to synthesize witness values that
  make the program go wrong.
  We prove that our procedure synthesizes
  {general} witnesses in that if a witness is
  found, then for all inhabited input types,
  there exist values that can make the function go wrong.
  %
  Second, we show how to extend the above procedure to
  produce a reduction graph that can be used to
  interactively visualize and debug witness executions.
  %
  Third, we evaluate the coverage of our approach
  on two data sets comprising over 4,500 ill-typed
  student programs.
  Our technique is able to generate witnesses for
  around 85\% of the programs, and our reduction graph
  yields small counterexamples for over 80\% of the witnesses.
  %
  Finally, we evaluate whether our witnesses help
  students understand and fix type errors, and
  find that students presented with our witnesses
  show a greater understanding of type errors
  than those presented with a standard error message.
\end{abstract}

### 2017-04-27 paper/jfp-submission/evaluation.tex
tweak numbers...

\begin{figure}[htp]
\centerline{
\begin{minipage}{1.2\textwidth}
\includegraphics[width=0.5\linewidth]{coverage.png}
% \end{minipage}
% \begin{minipage}{\linewidth}
\includegraphics[width=0.5\linewidth]{distrib.png}
\end{minipage}
}
\vspace{3ex}
\caption{Results of our coverage testing and the distribution of test
  outcomes. Our random search successfully finds witnesses for 75--82\% of
  the programs in under one second, improving to 84\% in under
  10 seconds. In both datasets we detect actual type errors at least 75\%
  of the time, unbound variables or constructors 4--5\% of the time, and
  diverging loops 3\% of the time. For the remaining 13--16\% of the
  programs we are unable to provide any useful feedback.  }
\label{fig:results-witness}
\end{figure}

\paragraph{Results}
\label{sec:results-witness}
The results of our experiments are summarized in
Figure~\ref{fig:results-witness}.
%
In both datasets our tool was able to find a witness for at least 75\% of the
programs in under one second, \ie\ fast enough to be integrated as a
compile-time check. If we extend our tolerance to a 10 second timeout,
we hit a maximum of 83--86\% coverage.
%
Interestingly, while the vast majority of witnesses corresponded to a
type-error, as expected, 4--5\% triggered an unbound variable error (even
though \ocaml\ reported a type error) and 3\% triggered an infinite
recursion error.
%
For the remaining 13--16\% of programs we were unable to provide any useful
feedback as they either completed 1,000 tests successfully, or timed out
after one minute.
%
% XX programs were deemed safe and XX timed out even at 3,000 steps, \ie
% we could not provide any useful feedback for XX\% of the total programs.
%
While a more advanced search procedure, \eg\ dynamic-symbolic execution,
could likely uncover more errors, our experiments suggest that
type errors are coarse enough (or that novice programs are \emph{simple}
enough) that these techniques are not necessary.

@@

% \item \ES{others?}
%
\begin{figure}[t]
\centerline{
\begin{minipage}{1.2\textwidth}
\includegraphics[width=0.5\linewidth]{trace_size_step.png}
\includegraphics[width=0.5\linewidth]{trace_size_jump.png}
\end{minipage}
}
\vspace{3ex}
\caption{Complexity of the generated traces. 81\% of the combined traces
  have a jump complexity of at most 10, with an average complexity of 7
  and a median of 5.}
\label{fig:results-complexity}
\end{figure}
%
\paragraph{Results}
\label{sec:results-complexity}
The results of the experiment are summarized in
Figure~\ref{fig:results-complexity}.
%
The average number of single-step reductions per trace is 31 for the
\ucsdbench\ dataset (35 for the \uwbench\ dataset) with a maximum of
2,745 (\resp 986) and a median of 17 (\resp 17).
%
The average number of jumps per trace is 7 (\resp 7) with a
maximium of 353 (\resp 185) and a median of 4 (\resp 4).
%
In both datasets 80\% or more traces have at most 10 jumps.

>>>>

\begin{figure}[htp]
\centerline{
\begin{minipage}{1.2\textwidth}
\includegraphics[width=0.5\linewidth]{coverage.png}
% \end{minipage}
% \begin{minipage}{\linewidth}
\includegraphics[width=0.5\linewidth]{distrib.png}
\end{minipage}
}
\vspace{3ex}
\caption{Results of our coverage testing and the distribution of test
  outcomes. Our random search successfully finds witnesses for 76--82\% of
  the programs in under one second, improving to 84--85\% in under
  10 seconds. In both datasets we detect actual type errors at least 75\%
  of the time, unbound variables or constructors 4--5\% of the time, and
  diverging loops 3\% of the time. For the remaining 13--16\% of the
  programs we are unable to provide any useful feedback.  }
\label{fig:results-witness}
\end{figure}

\paragraph{Results}
\label{sec:results-witness}
The results of our experiments are summarized in
Figure~\ref{fig:results-witness}.
%
In both datasets our tool was able to find a witness for over 75\% of the
programs in under one second, \ie\ fast enough to be integrated as a
compile-time check. If we extend our tolerance to a 10 second timeout,
we reach 84--85\% coverage, and if we allow a 60 second search,
we hit a maximum of 84--87\% coverage.
%
Interestingly, while the vast majority of witnesses corresponded to a
type-error, as expected, 4--5\% triggered an unbound variable error (even
though \ocaml\ reported a type error) and 3\% triggered an infinite
recursion error.
%
For the remaining 13--16\% of programs we were unable to provide any useful
feedback as they either completed 1,000 tests successfully, or timed out
after one minute.
%
% XX programs were deemed safe and XX timed out even at 3,000 steps, \ie
% we could not provide any useful feedback for XX\% of the total programs.
%
While a more advanced search procedure, \eg\ dynamic-symbolic execution,
could likely uncover more errors, our experiments suggest that
type errors are coarse enough (or that novice programs are \emph{simple}
enough) that these techniques are not necessary.

@@

% \item \ES{others?}
%
\begin{figure}[t]
\centerline{
\begin{minipage}{1.2\textwidth}
\includegraphics[width=0.5\linewidth]{trace_size_step.png}
\includegraphics[width=0.5\linewidth]{trace_size_jump.png}
\end{minipage}
}
\vspace{3ex}
\caption{Complexity of the generated traces. Over 80\% of the combined traces
  have a jump complexity of at most 10, with an average complexity of 7
  and a median of 5.}
\label{fig:results-complexity}
\end{figure}
%
\paragraph{Results}
\label{sec:results-complexity}
The results of the experiment are summarized in
Figure~\ref{fig:results-complexity}.
%
The average number of single-step reductions per trace is 30 for the
\ucsdbench\ dataset (45 for the \uwbench\ dataset) with a maximum of
2,745 (\resp 982) and a median of 14.5 (\resp 14.5).
%
The average number of jumps per trace is 7 (\resp 10) with a
maximium of 353 (\resp 221) and a median of 4 (\resp 4).
%
In both datasets 80\% or more traces have at most 10 jumps.

### 2017-04-27 paper/jfp-submission/introduction.tex
tweak numbers...

\paragraph{3. Evaluating Witnesses}
%
Of course, the problem of finding witnesses is
undecidable in general. In fact, due to the necessarily
conservative nature of static typing, there
may not even exist any witnesses for a given
ill-typed program.
%
Thus, our approach is a heuristic that is only useful
if it can find \emph{compact} witnesses for
\emph{real-world} programs.
%
Our third contribution is an extensive evaluation of our approach
on two different sets of ill-typed programs obtained by instrumenting
compilers used in beginner's classes~(\S~\ref{sec:evaluation}).
%
The first is the \uwbench\ data set~\cite{Lerner2007-dt}
comprising \uwsize\ ill-typed programs.
%
The second is a new \ucsdbench\ data set, comprising \ucsdsize\
ill-typed programs.
%
We show that for both data sets, our technique is able to generate
witnesses for nearly 90\% of the programs, in under a second in the
vast majority of cases.
%
Furthermore, we show that a simple interactive strategy yields
compact counterexample traces with at most 5 steps for 57\%
of the programs, and at most 10 steps for 81\% of the programs.

>>>>

\paragraph{3. Evaluating Witnesses}
%
Of course, the problem of finding witnesses is
undecidable in general. In fact, due to the necessarily
conservative nature of static typing, there
may not even exist any witnesses for a given
ill-typed program.
%
Thus, our approach is a heuristic that is only useful
if it can find \emph{compact} witnesses for
\emph{real-world} programs.
%
Our third contribution is an extensive evaluation of our approach
on two different sets of ill-typed programs obtained by instrumenting
compilers used in beginner's classes~(\S~\ref{sec:evaluation}).
%
The first is the \uwbench\ data set~\cite{Lerner2007-dt}
comprising \uwsize\ ill-typed programs.
%
The second is a new \ucsdbench\ data set, comprising \ucsdsize\
ill-typed programs.
%
We show that for both data sets, our technique is able to generate
witnesses for around 85\% of the programs, in under a second in the
vast majority of cases.
%
Furthermore, we show that a simple interactive strategy yields
compact counterexample traces with at most 5 steps for 60\%
of the programs, and at most 10 steps for over 80\% of the programs.
